---
title: "asca_tds_usage"
author: "Michele Ricci"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{asca_tds_usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tempR}
  %\VignetteDepends{tidyverse}
  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This function applies ASCA decomposition to a Temporal Dominant Sensation (TDS) dataset. It is required that the dataset is put in a long format. The decomposition applied is based on the assumption of a normal data distribution. Each decomposition is applied to each combination of units of time and sensory descriptors. 

### Wrangling procedure
To be used properly, the function needs to be applied to a `data.frame` object in a long format.
Furthermore, the columns referring to the factors included in the model must be of type factor, and the column referring to the analysis time must be numeric.


```{r}
library(tempR)
library(tidyverse)

data <- tempR::bars
data.long <- data %>% gather( time, CATA, 5:455) %>% mutate(time = str_remove(time, "time_") %>% str_remove(., "s$"))

```


Here it is a snippet of the optimal dataset structure.

```{r}
head(data.long)

str(data.long)
```


### Application of the function

Once the dataset is arranged in long format with a column for the time values and a column for the attribute values, we apply time-resolved ASCA decomposition on the dataset using the function asca_tds().

```{r}
library(ASCATCATA)


test_tds <- asca_tds(CATA~(sample+assessor)^2, data = data.long, timecol = "time", attributes = "attribute")
```


The resulting object is a list containing multiple objects:

* **sample**:A PCA object (as estimated by `prcomp` function) for a factor defined in the formula of the function.
* **assessor**: A PCA object (as estimated by `prcomp` function) for a factor defined in the formula of the function.
* **sample:assessor**: A PCA object (as estimated by `prcomp` function) for a factor defined in the formula of the function.
* **Residuals**: A PCA object estimated from the residual structure.
* **Info**: a list containing information about the sensory descriptors, the time units, and the parameter specified for the asca decomposition.
* **Parameters**: A data.frame containing all the residuals and fitted values estimated from the glm decomposition.






### plotting the results



The results can be reported using the plot_ASCA function. Calling the function with only the object obtained from the asca_tds() function will print one bi-plot for each factor plus one other bi-plot for residuals. It is possible to call one plot at a time specifying a number or a string containing the exact name of the factor as reported in the formula.


``` {r , fig.dim = c(8, 6)}
#plot_ASCA(test_tds) This will print all the plots one after the other.

plot_ASCA(test_tds, object = 1)
# plot_ASCA(test_tds, object = "sample") This has the same result

```



``` {r , fig.dim = c(8, 6) }

plot_ASCA(test_tds, object = 2)
# plot_ASCA(test_tds, object = "assessor") This has the same result

```


```{r , fig.dim = c(8, 6)}
plot_ASCA(test_tds, object = 3)
# plot_ASCA(test_tds, object = "sample:assessor") This has the same result
```



```{r , fig.dim = c(8, 6)}
plot_ASCA(test_tds, object = 4)
# plot_ASCA(test_tds, object = "Residuals") This has the same result
```

It is possible to select different arrangements for the depiction of loading values specifying the different parameters of the plot_ASCA() function. 
By defining a number or the name of the factor in **object** it is possible to select which plot will be printed. Modifying the values of **density**, **path**, **time.label**, and **path.smooth** parameters it is possible to modify the aesthetic of the plot.

```{r , fig.dim = c(8, 6)}
plot_ASCA(test_tds, object = 1, density = TRUE, path = FALSE, path.smooth = FALSE)
```

The values reported here are the same as the plot above. This depiction is suggested only if it is not necessary to report information about the time structure, but onl about the attributes. 
May be useful to summarize the differences due to the overall use of sensory descriptors.


### ASCA analysis with short loading structure 

```{r , fig.dim = c(8, 6)}
test_tds_short <- asca_tds(CATA~(sample+assessor)^2, data = data.long, timecol = "time", attributes = "attribute", loadings.time.structure = "short")

```



```{r , fig.dim = c(12, 6)}

plot_ASCA(test_tds_short)

```


### plotting loadings/contribution

It is possible to plot the contribution of each attribute on the overall variability for each factor by adopting the function plot_time_loadings()


```{r , fig.dim = c(8, 6)}

plot_time_loadings(test_tds)

```




