---
title: "asca_ti_usage"
author: "Michele Ricci"
date: "2023-11-25"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{asca_ti_usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tempR}
  %\VignetteDepends{tidyverse}
  %\VignetteDepends{readxl}
  %\VignetteDepends{httr}
  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This function applies ASCA decomposition to a Time-Intensity (TI) dataset. It is required that the dataset is put in a long format. The decomposition applied is based on the assumption of a normal data distribution. Each decomposition is applied to each unit of time.  


### Wrangling procedure

The wrangling procedure required to prepare a dataset for the application of the function is reported.
An open Time Intensity dataset is taken from the website https://help.xlstat.com/dataset/time-intensity-data_0.xlsm


```{r , warning=F, message=F}

library(readxl)
library(httr)
library(tidyverse)
library(ASCATCATA)

url1 <- "https://help.xlstat.com/dataset/time-intensity-data_0.xlsm"

GET(url1, write_disk(tf <- tempfile(fileext = ".xlsm")))

tf <- str_replace_all(tf, "\\\\", "//")
data <- read_excel(tf)
```

After downloading the dataset, it is necessary to arrange it in the required structure. It is necesessary to put it in the long format and to define a specific column containing the time data.

```{r}
data.long <- data %>% gather(time, intensity, 6:36) %>% filter(time != "0") %>% droplevels()

str(data.long)

```

Once the dataset has been prepared, the asca_ti() function is applied, defining the formula applied by the ASCA decomposition and the column containing the time column


``` {r}

test_ti <- asca_ti(intensity ~ PANELIST+PRODUCT, data = data.long, timecol = "time")
```


The resulting object is a list containing multiple objects:

* **PANELIST**:A PCA object (as estimated by `prcomp` function) for a factor defined in the formula of the function.
* **PRODUCT**: A PCA object (as estimated by `prcomp` function) for a factor defined in the formula of the function.
* **Residuals**: A PCA object estimated from the residual structure.
* **Info**: a list containing information about the sensory descriptors, the time units, and the parameter specified for the asca decomposition.
* **Parameters**: A data.frame containing all the residuals and fitted values estimated from the glm decomposition.

